Automatically Splitting a Staged Lambda Calculus

### 2015-07-10 paper/abstract.tex
various small edits
reviewer comments

Since our work builds on a lambda calculus, our splitting algorithm
enables the programmer to write staged programs by using powerful
programming abstractions in a composable and modular fashion. Our
splitting transformation maps such programs into multi-pass programs
that can run significantly more efficiently by assigning frequent
computations to earlier passes so that they can be reused (cheaply)
later.  Our experiments based on a prototype implementation show that
our splitting algorithm is able to perform highly non-trivial algorithmic
transformations, sometimes automatically improving efficiency
asymptotically by nearly a linear factor.  Since the splitting
transformation takes place at compile time, these benefits are
achieved without significant run-time penalty.

>>>>

Since our work builds on a lambda calculus, our splitting algorithm
enables the programmer to write staged programs by using powerful
programming abstractions in a composable and modular fashion. Our
splitting transformation maps such programs into multi-pass programs
that can run asymptotically more efficiently by assigning frequent
computations to earlier passes so that they can be reused (cheaply)
later.  Our experiments based on a prototype implementation show that
our splitting algorithm is able to perform highly non-trivial algorithmic
transformations, sometimes automatically improving efficiency
asymptotically by nearly a linear factor.  Since the splitting
transformation takes place at compile time, these benefits are
achieved without significant run-time penalty.

\TODO We also proved some stuff.

### 2015-07-10 paper/examples.tex
reviewer comments

Our algorithm, on the other hand, produces:
%
\begin{lstlisting} 
datatype binnat = Zero | Even of nat | Odd of nat

@@

Let us return to the quickselect algorithm, which we discussed at
length in \ref{sec:overview}. (The code is in \ref{fig:qs-staged}.)
\texttt{qss} finds the $k$th largest element of a list $l$ by
recursively \texttt{partition}ing the list by its first element, then recurring
on the side containing the $k$th largest element. $l$ is first-stage and $k$ is
second-stage.

@@

2`fun qs2 (((), k : int), p : tree) = 
  case p of
    Leaf => 0
  | Branch (n,left,h,right) =>
      case compare (k, n) of 
        LT => qs2 (((), k), left) 
      | EQ => h 
      | GT => qs2 (((), k-n-1), right)`
\end{lstlisting}
%
This is nearly identical to the cleaned-up code we presented in
\ref{fig:qs-split}, except that the mixed-stage argument to both passes 
is explicitly masked, and we do not suppress the trivial first-stage result
returned by \texttt{qs1}.

\texttt{qs1} \texttt{partition}s $l$, but since the comparison with $k$ (to
determine which half of $l$ to recur on) is at the second stage, it simply
recurs on \emph{both} halves, pairing up the results along with $h$ (the head of
$l$) and $n$ (the size of the left half). \texttt{qs2} takes $k$ and this tree
$p$, and uses $k$ to determine how to traverse $p$.

>>>>

Our algorithm, on the other hand, produces:
%
\begin{lstlisting} 
datatype binnat = Zero
                | Even of binnat 
                | Odd of binnat

@@

Let us return to the quickselect algorithm, which we discussed at
length in \ref{sec:overview}. (The code is in \ref{fig:qs-staged}.)
\texttt{qss} finds the $k$th largest element of a list $l$ by
recursively partitioning the list by its first element, then recurring
on the side containing the $k$th largest element. $l$ is first-stage and $k$ is
second-stage.

@@

2`fun qs2 (((), k : int), p : tree) = 
  case p of
    Leaf => 0
  | Branch (n,left,h,right) =>
      case compare k n of 
        LT => qs2 (((), k), left) 
      | EQ => h 
      | GT => qs2 (((), k-n-1), right)`
\end{lstlisting}
%
This is nearly identical to the cleaned-up code we presented in
\ref{fig:qs-split}, except we do not suppress the trivial inputs and outputs of 
\texttt{qs1} and \texttt{qs2}.

The function \texttt{qs1} partitions $l$, but since the comparison with $k$ (to
determine which half of $l$ to recur on) is at the second stage, it simply
recurs on \emph{both} halves, pairing up the results along with $h$ (the head of
$l$) and $n$ (the size of the left half). \texttt{qs2} takes $k$ and this tree
$p$, and uses $k$ to determine how to traverse $p$.

### 2015-07-10 paper/intro.tex
minor fixes
reviewer comments

Programs featuring this structure are particularly important in real-time
computer graphics; in fact, modern graphics architectures \emph{require} that
graphics computations be organized into distinct {\em passes} which perform
increasingly fine-grained computations (for example, per object, per region of
the screen, and per pixel, where many computations in later passes use the
results from a single computation in an earlier pass)\,\cite{OpenGL4Spec}.
While performant, this paradigm of programming with explicit passes results in
complex code where invariants must hold across different passes, and local
changes for one pass may require changes to other passes to ``plumb'' inputs
from all other frequencies\,\cite{Foley:2011}.
%
In other words, requiring manual separation of passes breaks key programming
abstractions such as composition and modularity.
%
Graphics researchers therefore have suggested using explicitly staged programs
\,\cite{Proudfoot:2001,Foley:2011,He:2014}, deferring to the compiler the
mechanical task of separating such programs into passes.

@@

While partial evaluation and metaprogramming have been studied extensively,
frequency reduction techniques have not seen the same attention. In
\cite{JS86-staging}, J{\o}rring and Scherlis proposed {\em pass separation} as a
manual method for {\em splitting} a program into multiple passes.
%
In our example, pass separation would split the function $f$ into two passes
$f_1$ and $f_2$ such that $f(x,y_i) = f_2(f_1(x),y_i)$. Then we evaluate $F$ by
evaluating the first pass on $x$, and then the second pass on each $y_i$.
%
\begin{lstlisting}
fun @$F_\textrm{multipass}(x, y_1, y_2, \ldots, y_m)$@ = 
  let @$z = f_1(x)$@
  in @$f_2(z, y_1);  f_2(z, y_2); \ldots; f_2(z, y_m)$@
\end{lstlisting}
%
The key difference between pass separation and partial evaluation (or
metaprogramming) is that the former can be performed without access to
the first argument $x$; $F_\textrm{multipass}$ works for any $x$,
while $F_v$ is defined only for $x=v$.  Note that unlike
precomputation, frequency reduction does not require knowledge of
arguments; however, if a particular $v$ is known, we can use partial evaluation
to specialize $F_\textrm{multipass}$ to that $v$.

>>>>

Programs featuring this structure are particularly important in real-time
computer graphics; in fact, modern graphics architectures \emph{require} that
graphics computations be organized into distinct {\em passes} which perform
increasingly fine-grained computations (for example, per object, per region of
the screen, and per pixel, where many computations in later passes use the
results from a single computation in an earlier pass)\,\cite{OpenGL4Spec}.
While performant, this paradigm of programming with explicit passes results in
complex code where invariants must hold across different passes, and local
changes for one pass may require changes to other passes to ``plumb'' inputs
from all other frequencies (\TODO explain what ``frequency'' means here)\,\cite{Foley:2011}.
%
In other words, requiring manual separation of passes breaks key programming
abstractions such as composition and modularity.
%
Graphics researchers therefore have suggested using explicitly staged programs
\,\cite{Proudfoot:2001,Foley:2011,He:2014}, deferring to the compiler the
mechanical task of separating such programs into passes.

@@

While partial evaluation and metaprogramming have been studied extensively,
frequency reduction techniques have not seen the same attention. In
\cite{JS86-staging}, J{\o}rring and Scherlis proposed {\em pass separation} as a
manual method for {\em splitting} a program into multiple passes.
%
In our example, pass separation would split the function $f$ into two passes
$f_1$ and $f_2$ such that $f(x,y_i) = f_2(f_1(x),y_i)$. Then we evaluate $F$ by
evaluating the first pass on $x$, and using the result $z$ to evaluate the
second pass on each $y_i$.
%
\begin{lstlisting}
fun @$F_\textrm{multipass}(x, y_1, y_2, \ldots, y_m)$@ = 
  let @$z = f_1(x)$@
  in @$f_2(z, y_1);  f_2(z, y_2); \ldots; f_2(z, y_m)$@
\end{lstlisting}
%
The key difference between pass separation and partial evaluation (or
metaprogramming) is that the former can be performed without access to
the first argument $x$; $F_\textrm{multipass}$ works for any $x$,
while $F_v$ is defined only for $x=v$.  Note that unlike
precomputation, frequency reduction does not require knowledge of
arguments; however, if a particular $v$ is known, we can use partial evaluation
to specialize $F_\textrm{multipass}$ to that $v$.

### 2015-07-10 paper/overview.tex
reviewer comments

Suppose that we wish to perform a series of order statistics queries
on a list \texttt{l}. To this end, we can use the quickselect
algorithm~\cite{Hoare:1961}, which given a list \texttt{l} and a rank
\texttt{k}, returns the element of \texttt{l} with rank \texttt{k}.
As implemented in an ML-like language in \ref{fig:qs-unstaged},
\texttt{qs} partitions \texttt{l} using the first element as
a pivot, and then recurs on one of the two resulting sides, depending on
the relationship of \texttt{k} to the size \texttt{i} of the first half, in
order to find the desired element.  Assuming that the input is uniformly
randomly ordered (which can be achieved by pre-permuting the input), \texttt{qs}
runs in expected $O(n)$ time.
%
Using \texttt{qs}, we can perform $m$ different order statistics queries with
ranks $\mathtt{k1},\dots,\mathtt{km}$ as follows:
%
\begin{lstlisting}
(qs l k1, qs l k2, ..., qs l km)
\end{lstlisting}
%
Unfortunately, this approach requires $O(n \cdot m)$ time.

@@

The process of adding staging annotations to unstaged code has been
the subject of extensive research (\ref{sec:related}). In this paper, we do not
consider this problem, instead assuming that the annotations have been provided.
In the case of \texttt{qss}, we have specifically chosen annotations which
maximize the work performed in the first stage.

@@

In the rest of this section, we present a high-level overview of the main ideas
behind our splitting algorithm, applied to \texttt{qss}.
%
Splitting \texttt{qss} yields a two-pass program that creates a
probabilistically balanced, augmented binary search tree (which is isomorphic to
a treap data structure~\cite{SeidelAr96}) as an intermediate data structure. In
particular, its first pass (\texttt{qs1}) constructs such a binary search tree,
and its second pass (\texttt{qs2}) traverses the tree, using the embedded size
information to find the element of the desired rank.
%
The code for \texttt{qs1} and \texttt{qs2} is in \ref{fig:qs-staged}.

>>>>

Suppose that we wish to perform a series of order statistics queries
on a list \texttt{l}. To this end, we can use the quickselect
algorithm~\cite{Hoare:1961}, which given a list \texttt{l} and a rank
\texttt{k}, returns the element of \texttt{l} with rank \texttt{k}.
As implemented in an ML-like language in \ref{fig:qs-unstaged},
\texttt{qs} partitions \texttt{l} using the first element as
a pivot, and then recurs on one of the two resulting sides, depending on
the relationship of \texttt{k} to the size \texttt{i} of the first half, in
order to find the desired element.  
If the index is out of range, a default value of \texttt{0} is returned.
Assuming that the input is uniformly
randomly ordered (which can be achieved by pre-permuting the input), \texttt{qs}
runs in expected $O(n)$ time.
%
Using \texttt{qs}, we can perform $m$ different order statistics queries with
ranks $\mathtt{k1},\dots,\mathtt{km}$ as follows:
%
\begin{lstlisting}
(qs l k1, qs l k2, ..., qs l km)
\end{lstlisting}
%
Unfortunately, this approach requires $O(n \cdot m)$ time.

@@

The process of automatically adding staging annotations to unstaged code has been
the subject of extensive research (\ref{sec:related}). In this paper, we do not
consider this problem, instead assuming that the annotations already exist.
In the case of \texttt{qss}, we have specifically chosen annotations which
maximize the work performed in the first stage.

@@

In the rest of this section, we present a high-level overview of the main ideas
behind our splitting algorithm, applied to \texttt{qss}.
%
Splitting \texttt{qss} yields a two-pass program that creates a
probabilistically balanced, augmented binary search tree (which is isomorphic to
a treap data structure~\cite{SeidelAr96}) as an intermediate data structure. In
particular, its first pass (\texttt{qs1}) constructs such a binary search tree,
and its second pass (\texttt{qs2}) traverses the tree, using the embedded size
information to find the element of the desired rank.
%
The code for \texttt{qs1} and \texttt{qs2} is in \ref{fig:qs-split}.

### 2015-07-10 paper/relatedwork.tex
reviewer comments

Early approaches to partial evaluation can be viewed as operating in
two stages: binding time analysis and program specialization.  For a
multivariate program with clearly marked static and dynamic arguments,
binding-time analysis identifies all the parts of the program that can
be computed by the knowledge of static arguments. Using this
information and the values of static arguments, program specialization
specializes the original program to a partially-evaluated one that
operates on many different dynamic arguments.  This approach has been
applied to construct partial evaluators for a number of languages such
as Scheme~\cite{OB91-Similix,Consel88-Schism}, although for the most
part restricted to first-order subsets.

>>>>

Early approaches to partial evaluation can be viewed as operating in
two stages: binding time analysis and program specialization.  For a
multivariate program with clearly marked static and dynamic arguments,
binding-time analysis identifies all the parts of the program that can
be computed by the knowledge of static arguments. Using this
information and the values of static arguments, program specialization
specializes the original program to a partially-evaluated one that
operates on many different dynamic arguments.  This approach has been
applied to construct partial evaluators for a number of languages such
as Scheme~\cite{OB91-Similix,Consel88-Schism}.

### 2015-07-10 paper/semantics.tex
reviewer comments

All three worlds contain unit, product, and function types defined in the usual fashion;
sum and recursive types are provided for worlds \bbonep\ and \bbtwo\ 
(we omit them from \bbonem, because they are not needed for our examples).
These ``standard'' features can only be constructed from subterms of the same world,
and variables can only be used at the same world where they were introduced.
Thus differing worlds (and hence, differing stages of computation) only interact by means
of the $\fut$ and $\curr$ type formers.  These modalities are internalizations
of typing at worlds \bbtwo\ and \bbonep, respectively, to typing at \bbonem.

>>>>

All three worlds contain unit, product, and function types defined in the usual fashion;
sum and recursive types are provided for worlds \bbonep\ and \bbtwo.
\footnote{We omit sums and recursive types from \bbonem\ to save space, though we do not believe they would present any unique difficulty.}
These ``standard'' features can only be constructed from subterms of the same world,
and variables can only be used at the same world where they were introduced.
Thus differing worlds (and hence, differing stages of computation) only interact by means
of the $\fut$ and $\curr$ type formers.  These modalities are internalizations
of typing at worlds \bbtwo\ and \bbonep, respectively, to typing at \bbonem.
